Functional Pearl: Short and Mechanized Logical Relation for Dependent Type Theories

### 2024-12-09 paper/paper.tex
Reorganizing Section 5 up to Adequacy

\begin{figure}
\drules[P]{$[[a => b]]$}{Parallel Reduction}{AppAbs, IfTrue, IfFalse, JRefl}
\caption{Parallel reduction ($\beta$ rules only) }
\label{fig:par}
\end{figure}

@@

\begin{figure}
  \[
    \begin{array}{llcl}
       \beta\text{-}\mathit{neutral\ terms} &
      [[e]] & ::= & [[x]]\ |\ [[e f]]\ |\ [[J e f]]\ |\ [[if e f
                    f]] \\ \\
      \beta\text{-}\mathit{normal\ terms} &
      [[f]] & ::= & [[e]]\ |\ [[Set i]]\ |\ [[Void]]\ |\ [[Pi x : f . f]]\
                    |\ [[f ~ f : f]]\\
            & & |   & [[\ x . f]]\ |\ [[refl]]\ |\ [[Bool]]\ |\ [[true]]\ |\ [[false]]
    \end{array}
  \]
  \caption{$\beta$-neutral and normal forms}
  \label{fig:nenf}
\end{figure}

@@

\begin{figure}
  \drules[I]{$[[Interp I i A S]]$}{Extended Logical Relation}{Ne, VoidNew, BoolNew, EqNew}
  \caption{Extended logical relation for \lang{} (new and changed rules)}
  \label{fig:logrelopen}
\end{figure}

@@

@@

Because we are working with open terms, we need a few additional syntactic
lemmas about reduction. First, a renaming $[[xi]]$ is a generalization of
weakening when working with simultaneous substitutions. It consistently maps the
variables that appears in terms to other variables. If a renamed term has been reduced, we can
always recover the result of the reduction without the renaming.

@@

We can show that parallel reduction preserves $\beta$-normal and
neutral forms.
\begin{lemma}[Par preserves $\beta$-neutral and normal forms\footnote{\dotv{normalform.v}{nf\_ne\_preservation}}]
  \label{lemma:parnenf}
  If $[[a => b]]$, then
  \begin{itemize}
  \item $[[ne a]]$ implies $[[ne b]]$
  \item $[[nf a]]$ implies $[[nf b]]$
  \end{itemize}
\end{lemma}
Lemma~\ref{lemma:parnenf} could have been strengthened to say that if
$[[ne a]]$ or $[[nf a]]$ and $[[a => b]]$, then $[[a = b]]$. Since
$\ottkw{ne}$ and $\ottkw{nf}$ captures terms free of $\beta$ redexes, parallel
reduction cannot take any real reduction steps and can only step into a term
itself. However, for the purpose of our proof, Lemma~\ref{lemma:parnenf} is
sufficient.

All the properties we have shown in
Section~\ref{sec:logreldep} and \ref{sec:logrelproof} before the fundamental
lemma can be proven in the same order, where the new cases due to \rref{I-Ne}
and the augmentation of neutral terms to \rref{I-Void, I-Eq, I-Bool} can be
immediately discharged by Lemma~\ref{lemma:parnenf}.

Furthermore, Lemma~\ref{lemma:parnenf}, in its current weaker form, would
still hold after we extend our equational theory with the function $\eta$
rule, where parallel reduction can take $\eta$ steps but still preserves
$\beta$-normal form.

We also need to know that the $[[wne a]]$ and $[[wn a]]$ relations can be
justified compositionally. For example, an application has a neutral form when
the function has a neutral form and the argument has a normal form.

\begin{lemma}[Wne application\footnote{\dotv{normalform.v}{wne\_app}}]
  \label{lemma:wnewn}
  If $[[wne a]]$ and $[[wn b]]$, then $[[wne (a b)]]$.
\end{lemma}
\begin{proof}
  Immediate by induction over the length of the reduction sequences in
  $[[wne a]]$ and $[[wn b]]$.
\end{proof}

Furthermore, if we know that an application of a term to a variable
has a normal form, then we know that the term must have a normal form.
\begin{lemma}[Wn extensionality\footnote{\dotv{normalform.v}{ext\_wn}}]
  \label{lemma:extwn}
  If $[[wn (a x)]]$, then $[[wn a]]$.
\end{lemma}
\begin{proof}
  By induction over the length of the reduction sequence in $[[wn (a
  x)]]$. The conclusion follows from Lemmas~\ref{lemma:parantirenaming} and
  \ref{lemma:parnenf}.
\end{proof}

@@

  The \rref{I-Set} case is the most interesting. We must show that
  for all $[[j < i]]$, the set $[[{ A | exists S, InterpR j A S  }]]
  \in CR$. We immediately know that $[[{A | exists S, InterpR j A S
  }]] \in CR_1$ by \rref{I-Ne}. It remains to show that $[[{A | exists S, InterpR j A S
  }]] \in CR_2$, or equivalently, for all $[[A]]$, $[[InterpR j A S]]$
  implies $[[wn A]]$. Suppose $[[InterpR j A S]]$ for an arbitrary
  $[[A]]$. We have $[[InterpR j A S]]=[[Interp I j A S]]$ where $[[I]]$
  has the same definition from earlier but its domain restricted to
  numbers less than $[[j]]$. We perform another induction on
  the derivation of $[[Interp I j A S]]$. All cases are trivial except for the case for
  \rref{I-Pi}. Our induction hypothesis immediately gives us $[[wn
  A]]$. To derive $[[wn (Pi x : A . B)]]$, it remains to show $[[wn B]]$. We
  use the outermost induction hypothesis to show that $[[x]]$
  semantically inhabits $[[A]]$, from which we derive $[[wn (B {x / x})]]$
  and conclude $[[wn B]]$ through antirenaming (Lemma~\ref{lemma:parantirenaming}).
\end{proof}

>>>>

\begin{figure}
\drules[P]{$[[a => b]]$}{Parallel reduction}{AppAbs, IfTrue, IfFalse, JRefl}
\caption{Parallel reduction ($\beta$ rules only) }
\label{fig:par}
\end{figure}

@@

\begin{figure}
  \begin{align*}
    [[e]] & ::= [[x]] \mid [[e f]] \mid [[J e f]] \mid [[if e f f]] \mid [[absurd e]]
    && \textit{Neutral terms} \\
    [[f]] & ::= [[e]] \mid [[Set i]] \mid [[Pi x : f . f]] \mid [[Bool]] \mid [[f ~ f : f]] \mid [[Void]] \mid [[\ x . f]] \mid [[true]] \mid [[false]] \mid [[refl]]
    && \textit{Normal terms}
  \end{align*}
  \caption{$\beta$-neutral and $\beta$-normal forms}
  \label{fig:nenf}
\end{figure}

@@

Weakly normalizing terms are composed of other weakly normalizing terms
following the structure of neutral and normal forms. For example, an
application normalizes to neutral form if the function normalizes to neutral
form and the argument to normal form. Similar lemmas hold for
function types\footnote{\dotv{normalform.v}{wn\_pi}},
abstractions\footnote{\dotv{normalform.v}{wn\_abs}},
conditionals\footnote{\dotv{normalform.v}{wne\_if}},
and absurdity\footnote{\dotv{normalform.v}{wne\_absurd}}.

\begin{lemma}[Application (wne)\footnote{\dotv{normalform.v}{wne\_app}}]
  \label{lemma:wnewn}
  If $[[wne a]]$ and $[[wn b]]$, then $[[wne (a b)]]$.
\end{lemma}

\begin{proof}
  By induction over the length of reduction sequences in
  $[[wne a]]$ and $[[wn b]]$.
\end{proof}

@@

\begin{figure}
  \drules[I]{$[[Interp I i A S]]$}{Extended logical relation}{Ne, VoidNew, BoolNew, EqNew}
  \caption{Extended logical relation for \lang{} (new and changed rules)}
  \label{fig:logrelopen}
\end{figure}

@@

Because we are working with open terms, we need additional lemmas for parallel
reduction. First, it preserves neutral and normal forms.

\begin{lemma}[Preservation of neutral and normal forms (p.r.)\footnote{\dotv{normalform.v}{nf\_ne\_preservation}}]
  \label{lemma:parnenf}
  If $[[a => b]]$, then $[[ne a]]$ implies $[[ne b]]$,
  and $[[nf a]]$ implies $[[nf b]]$.
\end{lemma}

Since neutral and normal forms have no redexes, if $[[a]]$ is neutral or normal,
parallel reduction never $\beta$-reduces, so $[[a = b]]$ additionally holds,
but preservation of neutral and normal forms is sufficient for our proof.

Next, parallel reduction also satisfies an antirenaming property. A
\emph{renaming} $[[xi]]$ is a special case of substitution where all variables
are replaced by other variables. In this special case, if a renamed term
reduces, then we can separate the reduction from the renaming, asserting that
renaming never introduces new redexes.

@@

Together, these lemmas prove an inversion property for weakly normalizing forms:
if an application of a term to a variable is weakly normalizing, then the term
must also be weakly normalizing.

\begin{lemma}\hspace{-0.5em}\footnote{\dotv{normalform.v}{ext\_wn}}
  \label{lemma:extwn}
  If $[[wn (a x)]]$, then $[[wn a]]$.
\end{lemma}
\begin{proof}
  By induction over the length of the reduction sequence in $[[wn (a x)]]$.
  When the reduction is a $\beta$-reduction, $[[a]]$ is a function and the
  application reduces to a renaming by $[[x]]$, so the conclusion follows
  from Lemma~\ref{lemma:parantirenaming}.
\end{proof}

All the properties we have shown in
Section~\ref{sec:logreldep} and \ref{sec:logrelproof} before the fundamental
lemma can be proven in the same order, where the new cases due to \rref{I-Ne}
and the modifications to \rref{I-Void, I-Eq, I-Bool} can be discharged by
Lemma~\ref{lemma:parnenf}.

% \jc{Don't mention this right now; delay to $\eta$ section.}
% Furthermore, Lemma~\ref{lemma:parnenf}, in its current weaker form, would
% still hold after we extend our equational theory with the function $\eta$
% rule, where parallel reduction can take $\eta$ steps but still preserves
% $\beta$-normal form.

@@

  The \rref{I-Set} case is the most interesting. We must show that
  for all $[[j < i]]$, the set $[[{ A | exists S, InterpR j A S  }]]
  \in CR$. We immediately know that $[[{A | exists S, InterpR j A S
  }]] \in CR_1$ by \rref{I-Ne}. It remains to show that $[[{A | exists S, InterpR j A S
  }]] \in CR_2$, or equivalently, for all $[[A]]$, $[[InterpR j A S]]$
  implies $[[wn A]]$. Suppose $[[InterpR j A S]]$ for an arbitrary
  $[[A]]$. We have $[[InterpR j A S]]=[[Interp I j A S]]$ where $[[I]]$
  has the same definition from earlier but its domain restricted to
  numbers less than $[[j]]$. We perform another induction on
  the derivation of $[[Interp I j A S]]$. All cases are trivial except for the case for
  \rref{I-Pi}. Our induction hypothesis immediately gives us $[[wn
  A]]$. To derive $[[wn (Pi x : A . B)]]$, it remains to show $[[wn B]]$. We
  use the outermost induction hypothesis to show that an arbitrary $[[y]]$
  semantically inhabits $[[A]]$, from which we derive $[[wn (B {y / x})]]$
  and conclude $[[wn B]]$ through antirenaming (Lemma~\ref{lemma:parantirenaming}).
\end{proof}
